Contrasting patterns of richness, abundance, and turnover in mountain bumble bees and their floral hosts

### 2021-11-05 code/ms2_nullmodels.Rmd
updating just in case

ggsave("../ms_2/figures/null_model_ktype_preference.pdf", null_model_overall, height = 6, width = 7)
ggsave("../ms_2/figures/null_model_ktype_preference.png", null_model_overall, height = 6, width = 7)

@@

ggsave("../ms_2/figures/fig3.pdf", null_model_pbl_GAM, height = 6, width = 7)
ggsave("../ms_2/figures/null_model_ktypeXtonguelength_preference.png", null_model_pbl_GAM, height = 6, width = 7)

>>>>

ggsave("../ms_2/figures/fig2.pdf", null_model_overall, height = 4.5, width = 6)
ggsave("../ms_2/figures/fig2.png", null_model_overall, height = 4.5, width = 6)

@@

ggsave("../ms_2/figures/fig3.pdf", null_model_pbl_GAM, height = 6, width = 7)
ggsave("../ms_2/figures/fig3.png", null_model_pbl_GAM, height = 6, width = 7)

### 2021-11-05 ms_1/ms_revision1.Rmd
updating just in case

---
title: "The elevational structure of a mountain mutualism: richness, abundance, and β-diversity of bumble bees and their floral hosts"

author: Douglas B. Sponsler^1^*, Fabrice Requier^2^, Katharina Kallnik^1^, Alice Claßen^1^, A. Fabienne Maihoff^1^, Johanna Sieger^3^, Ingolf Steffan-Dewenter^1^

# date: `r Sys.Date()`

output:
  bookdown::word_document2:
    fig_caption: yes
    fig_width: 7
    fig_height: 7
    df_print: kable
    keep_md: yes
    number_sections: no
    toc: no
    reference_docx: Word_template.docx

bibliography: 
  - references.bib
#  - knitcitations.bib
csl: ecology.csl
# csl: https://www.zotero.org/styles/ecology

---

@@

Data are provided as private-for-peer review. This study also uses novel code. All data and code required to reproduce our analyses are archived on Dryad: https://datadryad.org/stash/share/M_4Yt9VVXTRm36FSfwwGzUs2OTXP9Fv52jjETdSu6ys

@@

Environmental gradients generate and maintain biodiversity on Earth. Mountain slopes are among the most pronounced terrestrial environmental gradients, and the elevational structure of species and their interactions can provide unique insight into the processes that govern community assembly and function in mountain ecosystems. We recorded bumble- bee-flower interactions over three years along an 1400 m elevational gradient in the German Alps. Using nonlinear modeling, we analyzed the elevational patterns at the nested levels of species abundance, species β-diversity, and interaction β-diversity. While floral richness exhibited mid-elevation peak, bumble bee richness increased with elevation before leveling off at the highest sites, demonstrating the exceptional adaptation of these bees to cold temperatures and short growing seasons. At a species level, though, bumble bees exhibited divergent responses to elevation in terms of abundance, with a clear separation between species preferring low vs. high elevations. Overall interaction β-diversity was dominated by strong turnover in the floral community, which exhibited well-defined threshold of β-diversity rate at the tree line ecotone. Turnover of bumble bees along the elevation gradient was modest, with the highest rate of β-diversity occurring over the interval from low- to mid-elevation sites. Interaction β-diversity increased sharply at the upper extreme of the elevation gradient (1800-2000 m), an interval over which we also saw steep decline in floral richness and abundance, suggesting a zone of stringent environmental filtering. Ongoing changes in land use are likely to cause reforestation in resource-rich mid-elevation meadows, forcing bumble bees into agriculturally intensified lowlands or resource-poor highlands, and these effects will interact with climate-induced range shifts in bumble bees and their floral hosts. The contrast between the relative stability of bumble bees community and the dynamism of the floral community in the Berchtesgadener Alps suggests that the strongest effects of climate and land use change on mountain bumble bees will be indirect effects mediated by the responses of their floral hosts.

*Keywords*: climate, beta-diversity, elevation gradient, *Bombus*, bumble bee, mountain ecology, pollination network

@@

Bumble bees (Hymenoptera: *Bombus* spp.) and their floral hosts provide an excellent system in which to study elevational patterns at both the species- and interaction-level. Bumble bees are believed to have arisen in the mountains of Asia during the cooling climate of the Eocene-Oligocene boundary, and their subsequent spread and diversification along mountain corridors and into lowland habitats was likely driven by alternating range expansion during cooling periods and retreat to higher elevations during warming periods [@Hines2008-er; @Stewart2010-jn; @Martinet2018-rb]. Today, they remain characteristically cold-adapted organisms [@Heinrich1994-cv], with peak abundance and diversity in mountain ranges and northern latitudes [@Williams1998-xi]. In these habitats, they are often the dominant guild of flower-visiting insects and the principal pollinators of entomophilous flora [@Goulson2010-lo]. As in the past, so now are mountain habitats critical theaters of ecological change for bumble bees. Warming climates are expected to cause habitat loss and potentially novel competitive pressures [@Kerr2015-dg; @Soroye2020-zb; though see @Guzman2021-my]. At the same time, socioeconomic drivers are causing the reforestation of pastures on mountain slopes and the intensification of agriculture in lowlands, both of which entail a loss of floral resources for bumble bees [@Tasser2002-om; @Tappeiner2006-jp; @Tasser2007-qo; @Walcher2017-kw].

In this paper, we present a study based on ~13,000 interactions between bumble bees (Hymenoptera: *Bombus*) and their floral hosts recorded along a 1400 m elevation gradient, complemented by corresponding surveys of floral abundance. Using this dataset, we investigate elevational responses of richness, abundance, and β-diversity, both at the level of species (bumble bees and flowering plants, respectively) and at the level of interactions (bumble- bee-flower pairs). Our results reveal nonlinear and, in some cases, contrasting elevational patterns, underscoring the fact that mountain slopes are complex ecological systems where multiple processes interact to produce biological responses. These patterns shed light on the basic ecology of the mountain mutualism between bumble bees and wildflowers, and they suggest how this mutualism might be reshaped by ongoing processes of climate and land use change.

@@

Sampling in each year of the study consisted of repeated visits to each site at approximately weekly intervals. Samples were only collected during periods without rain when the air temperature was at least 6°C. During each visit, interactions between bumble bee and flowers were recorded during a 50-minute transect walk. Bumblebees observed on a given flower were counted as floral visitors. Bumble bee queens were identified to species level in the field, while workers and males were stored in individually labeled tubes in the freezer for later identification in the laboratory according to [@Amiet1996-jz]. Floral visitation by male bumble bees was recorded during visitation sampling, but we chose to analyze only visitation by queens and workers. Males are often found resting on flowers, particularly late in the year, so their presence on a flower cannot necessarily be interpreted as a foraging visit. In conjunction with visitation observations, we estimated the flower cover of each herbaceous or shrubby plant species within each 60 x 60 m study plot to the nearest 0.01 m^2^. Plant species identification followed @Lauber2007-ha and @Oberndorfer2001-ro.

@@

We quantified bumble bee abundance as the total number of recorded floral visits per bumble bee species per site-date. We treated the cryptic *Bombus terrestris/lucorum* species pair as a single morphospecies, since these two species cannot be visually distinguished, and we pooled representatives of the parasitic *Psithyrus* subgenus --- *Bombus barbutellus*, *B. bohemicus*, *B. campestris*, *B. flavidus*, *B. quadricolor*, and *B. sylvestris* --- into the species group *B. psithryus*.  *Bombus hypnorum*, *B. humilis*, and *B. lapidarius* were extremely rare in our samples, so we omitted them from abundance analyses but retained them for β-diversity analyses.

@@

To facilitate clear interpretation with respect to our focal questions, the models above focus on elevational patterns of abundance by summarizing across seasonal time. For those interested in within-season abundance patterns and their interaction with elevation, more complex tensor-product GAMs are included in Appendix 1.

### Species and interaction β-diversity

For our analyses of species and interaction β-diversity, we pooled all samples by site irrespective of year. After pooling, we calculated β-diversity between all pairs of sites for bumble bees, flowering plants, and theirs interaction, respectively, using the binary Jaccard index. We then partitioned total interaction β-diversity into its components of species turnover (changes in species composition) and interaction rewiring (interaction turnover despite species constancy) [@Novotny2009-my; @Poisot2012-fk].

As an initial analysis, we visualized the relationship between site-wise elevation difference and each metric of β-diversity, fitting each response with a binomial regression smooth. To investigate the relationship between elevation and β-diversity more deeply, we performed a second analysis using generalized dissimilarity modeling (GDM) [@Ferrier2007-ji] to analyze the elevational variation in the β-diversity of bumble bees, flora, and their (unpartitioned) interactions, respectively. In addition to estimating the relationship between differences in a response variable and differences in a predictor variable, GDM captures the slope of this relationship over the range of the predictor variable, revealing potential variation in the amount of change in the response induced by a given change in the predictor [@Ferrier2007-ji]. This enabled us to ask whether the relationship between β-diversity and elevation exhibits thresholds or other nonlinearities. To control for potential confounding effects of sampling intensity, sampling days per site was included as a covariate. We also included the geographic distances between sites as a covariate to account for potential spatial autocorrelation. The importance of the variables in each model were assessed by a permutation test [@Fitzpatrick-2021aa].

@@

All analyses were conducted in `R` [@RCoreTeam-2021aa]. Data handling and visualization were performed with the `tidyverse` suite [@Wickham2019-aa]. GAM analyses were performed with packages `mgcv` [@Wood_2017], `mgcViz` [@Fasiolo2018aa], and `gratia` [@Simpson2021-aa]. β-diversity was calculated with the package `vegan` [@Oksanen_2019] and visualized using the packages `ggplot2` [@Wickham2016-aa] and `ggvegan` [@Simpson2019aa]. Calculation and partitioning of interaction β-diversity were performed with the package `bipartite` [@Dormann2008-aa; Dormann2009-aa]. GDM analysis of β-diversity was performed with the package `gdm` [@Fitzpatrick-2021aa]. Data and code to reproduce these analyses are archived on Dryad: https://datadryad.org/stash/share/M_4Yt9VVXTRm36FSfwwGzUs2OTXP9Fv52jjETdSu6ys.

@@

**Check to make sure this summary is still true**

## Summary of visitation and floral survey data

@@

The response of bumble bee richness to elevation exhibited an approximately linear increase up to 1600-1700 m and then leveled off and declined slightly at the upper limit of the gradient **(Figure 3A)**. Neither mowing nor abandonment had a significant effect on abundance relative to the reference management level of grazing (p > 0.05), but richness was significantly lower in 2012 than in the reference year of 2010 (p = 0.04) **(Table 1)**. In contrast, the response of floral richness exhibited a clear unimodal pattern, peaking between 1300-1500 m **(Figure 3B)**. Management had no significant effect on floral richness (P > 0.1), but floral richness was significantly lower in 2011 (p < 0.001) and 2012 (p = 0.03) than in 2010. The response of interaction richness to elevation was intermediate to that of bumble bee and floral richness, peaking around 1500 m and leveling off at higher elevation **(Figure 3C)**. Management did not have a significant effect on interaction richness (p > 0.1), but interaction richness was significantly higher in 2011 (p < 0.05) and lower in 2012 (p < 0.001) relative to the reference level of 2010.

Cumulative bumble bee abundance did not respond significantly to elevation (p > 0.1) or management (p > 0.1), but bumble bee abundance was significantly higher in 2011 than in the reference year of 2010 (p = 0.03) **(Figure 4A; Table 2)**. Floral abundance, however, peaked around 1000 and exhibited an approximately quadratic decline with increasing elevation (p = 0.003). Floral abundance was higher in abandoned meadows than in grazed meadows (p = 0.002) and lower in 2012 than in 2010 (p = 0.02).

@@

## Species and interaction β-diversity

Floral β-diversity among our study sites was very high overall and responded steeply to elevation difference between sites **(Figure 3A)**. Even sites at similar elevation exhibited ~40% species turnover, and the most vertically distant sites (elevation difference > 1250 m) differed by more than 85%. Bumble bee β-diversity was, in comparison to floral β-diversity, both lower overall and less responsive to elevational difference between sites **(Figure 3A)**. Sites at similar elevation exhibited ~25% species turnover, and species turnover between the most vertically distant sites remained less than 50%.

Total interaction β-diversity was >75% between sites at similar elevation and approached perfect dissimilarity in the most vertically distant sites **(Figure 3B)**. Partitioning revealed that total interaction β-diversity was driven primarily by species turnover, which accounted for ~50% of total β-diversity for sites at similar elevation and ~90% of total β-diversity for the most vertically distant sites. Species turnover consisted mainly of turnover in the floral community **(Figure 3C)**, though the joint turnover of plants and bumble bees accounted for ~25% of total species turnover between the most vertically distant sites. Interaction rewiring was most significant between sites at similar elevation, where it accounted for ~30% of total β-diversity, but its share of total β-diversity declined to ~10% in the most vertically distant sites **(Figure 3B)**.

Bumble bees, flora, and their interactions exhibited marked differences in their rates of β-diversity in response to elevation. **(Figure 4)**. Bumble bee turnover was relatively high from ~600-1100 m, briefly leveled off, and then accelerated again around 1250 m. Floral Turnover was high from ~600-1000 m, gradually leveled off from ~1000-1500 m, and then accelerated again around the tree line ecotone at ~1500 m. Interaction turnover mirrored the high rates of bumble bee and floral β-diversity from ~600-1000 m, but it exhibited a markedly different pattern above 1000 m, with relatively low turnover between 1000-1750 m and then a sharp acceleration from ~1750-2000 m. The conditional effect of elevation was significant (p < 0.001) in all models, and its importance was 4-20 time higher that of sampling intensity and geographic distance **(Table 4)**. Nevertheless, geographic distance was a significant predictor in all models (p < 0.001), and sampling intensity was a significant predictor in the bumble bee (p = 0.004) and interaction models (p < 0.001). Our GDMs explained 31% of total deviance in bumble bee β-diversity, 73% of floral β-diversity, and 71% of interactions β-diversity.

@@

Our study revealed strong and sometimes contrasting patterns of elevational structure in the richness, abundance, and β-diversity of mountain bumble bees and wildflowers. By extending our analysis to the level of species interactions, we show how the elevational patterns within bumble bee and wildflower communities are propagated to the functional relationships between these two guilds. These patterns suggest underlying processes governing the assembly of bumble- bee-wildflower networks in mountain habitats, shedding light on how this mutualism has evolved in the past and how it may respond to the nascent forces of climate and land use change.

@@

The relative stability of the bumble bee community in our study stands in stark contrast to the extreme dynamism of the floral community, as illustrated by the dominance of total interaction β-diversity by floral species turnover. This pattern is consistent with the findings of @Simanonok2014-ou in the Rocky Mountains of North America, and it is perhaps unsurprising given the exceptional cold-hardiness of bumble bees and the sensitivity of plants to temperature gradients. Nevertheless, it highlights the behavioral challenge that bumble bees have presumably faced for the entirety of their evolutionary history as mountain florivores: how to forage efficiently in a diverse floral community that exhibits rapid spatial and temporal turnover. A fascinating question that could not be answered by our sampling approach is whether mountain bumble bees actively exploit the elevational sensitivity of floral communities by adaptively foraging up- or downslope, as suggested by @Lundberg1980-qk. Bumble bees have large foraging ranges and have been shown both to track resources through space and time [@Devoto2014-ay] and to cross forest matrix to reach patches of foraging habitat [@Mola2020-yc]. Assuming an average slope of 20° and a foraging range of 1 km, a bumble bee could travel up- or downslope by more than 340 m, thus spanning an elevation band nearly 700 m wide. Such 3-dimensional foraging would enable bumble bees both to exploit the elevational turnover of floral species and to track preferred species through their elevationally staggered phenology, the latter constituting a sort of physiological time-travel [@Van_Straalen1983-df].

The roughly sigmoidal pattern of bumble bee β-diversity rate in response to elevation, revealed by our GDM analysis, probably reflects both the reciprocal abundance patterns of high- and low-elevation species and the occurrence of certain rarer species only at middle-elevation sites. As discussed above, though, bumble bee turnover was low over all, demonstrating the robustness of bumble bees to the environmental variation encompassed by our elevation gradient. Floral β-diversity rate also exhibited a sigmoidal pattern, and its inflection point coincides with the tree line ecotone at ~1500 m. The floristic importance of the tree line has been noted in previous work [e.g. @Pellissier2010-kg; @Descombes2017-id], and our study corroborates these findings specifically for the subset of flora visited by bumble bees. Interestingly, the tree line did not constitute a discrete transition to a new equilibrium of species composition, but rather a threshold marking the beginning of a process of β-diversity acceleration that continued for the remainder of the elevation gradient. While the most obvious effect of the tree line may be the relatively discrete transition from forest to scrub/grassland, the tree line also initiates a process of angiosperm species attrition that proceeds at a rate of about 40 species per 100 m [@Korner2003-sc], and the accelerated β-diversity we observed above the tree line mirrors simultaneous patterns of decreasing floral richness and abundance **(Figures 3B, 4B)**.

In contrast, the rate of interaction β-diversity exhibited a late and very pronounced acceleration at ~1800 m, indicating that the two highest sites at 1984 m and 2032 m differed strongly in interaction composition from the sites ~200 m below them. At these highest sites, the growing season is very short and vegetation is sparse, giving way to bare rock and scree landscapes, and it may be that the last 200 m interval of our elevation gradient imposes the most stringent environmental filter. It is interesting that interaction β-diversity did not exhibit the inflection point at the tree line ecotone (~1500 m) that we saw for floral β-diversity, despite the fact that overall interaction β-diversity was driven primarily by the turnover of the floral community. A possible explanation for this pattern is that there could be an elevational lag in bumble bee floral selection such that most bumble bee species do not switch to supra-tree-line floral species when these species first appear, but rather maintain associations with flora whose ranges span the tree line until the extreme species turnover between 1800 and 2000 m forces them to adopt new floral hosts.     
The poverty of floral resources at the highest elevations raises the question of why bumble bees choose to forage at such elevations at all. For some species, the selection of high elevation habitats may be influenced by thermal tolerances [@Oyen2016-ng], but that fails to explain the presence of elevation-generalist species (including the majority of the species in our study system) in resource-poor high-elevation habitats. It is possible that nest site limitation at lower elevations drives some bumble bees to establish nests at high elevation, and their ability to forage downslope is subsequently constrained by their flight range, which may be reduced due to the energetic costs of flight in cold environments [@Kovac2015-wy]. It is also possible that high elevation landscapes, while sparse in absolute floral resources, offer an escape from more competitive conditions at lower elevations where forager density is higher [@Hoiss2015-gv; @Classen2020-nh]. Similarly, higher elevations could also offer an escape from density-dependent parasite and pathogen pressure [@Parsche2018-ar]. Finally, it is worth speculating that an affinity for open, high-elevation habitats might be an evolutionary legacy in mountain bumble bees, since the existence of flower-rich meadows in the forest matrix below the tree line is largely an artifact of human land use arising mainly in the last 500 years of bumble bees' evolutionary history [@Hejcman2013-ca].

>>>>

---
title: "Contrasting patterns of richness, abundance, and turnover in mountain bumble bees and their floral hosts"

author: 
  - Douglas B. Sponsler^1^*  
  - Fabrice Requier^2^
  - Katharina Kallnik^1^
  - Alice Classen^1^
  - A. Fabienne Maihoff^1^
  - Johanna Sieger^3^
  - Ingolf Steffan-Dewenter^1^

output: 
  pdf_document: 
    keep_tex: yes
    latex_engine: xelatex 
    includes:
      in_header: preamble.tex

bibliography: 
  - references.bib
csl: ecology.csl
# csl: https://www.zotero.org/styles/ecology
---

\fontsize{12}{22}
\selectfont

@@

This study also uses novel code. All data and code required to reproduce our analyses are archived on Dryad: https://doi.org/10.5061/dryad.s1rn8pk7f

# Running Head: 
Mountain plants and pollinators

@@

Environmental gradients generate and maintain biodiversity on Earth. Mountain slopes are among the most pronounced terrestrial environmental gradients, and the elevational structure of species and their interactions can provide unique insight into the processes that govern community assembly and function in mountain ecosystems. We recorded bumble- bee-flower interactions over three years along an 1400 m elevational gradient in the German Alps. Using nonlinear modeling, we analyzed the elevational patterns at the nested levels of species abundance, species $\beta$-diversity, and interaction $\beta$-diversity. While floral richness exhibited mid-elevation peak, bumble bee richness increased with elevation before leveling off at the highest sites, demonstrating the exceptional adaptation of these bees to cold temperatures and short growing seasons. In terms of abundance, though, bumble bees exhibited divergent species-level responses to elevation, with a clear separation between species preferring low vs. high elevations. Overall interaction $\beta$-diversity was dominated by strong turnover in the floral community, which exhibited well-defined threshold of $\beta$-diversity rate at the tree line ecotone. Interaction $\beta$-diversity increased sharply at the upper extreme of the elevation gradient (1800-2000 m), an interval over which we also saw steep decline in floral richness and abundance. Turnover of bumble bees along the elevation gradient was modest, with the highest rate of $\beta$-diversity occurring over the interval from low- to mid-elevation sites. The contrast between the relative robustness bumble bees and sensitivity of plants to the elevational gradient in our study suggests that the strongest effects of climate change on mountain bumble bees will be indirect effects mediated by the responses of their floral hosts. Similarly, ongoing changes in land use are likely to cause succession to forest in resource-rich mid-elevation meadows, which may force bumble bees to seek floral resources in agriculturally intensified lowlands or the sparse highlands above the tree line.

**Keywords**: alpine plants, $\beta$-diversity, *Bombus*, climate,  elevation gradient, mountain ecology, pollination network

@@

Bumble bees (Hymenoptera: *Bombus* spp.) and their floral hosts provide an excellent system in which to study elevational patterns at both the species- and interaction-level. Bumble bees are believed to have arisen in the mountains of Asia during the cooling climate of the Eocene-Oligocene boundary, and their subsequent spread and diversification along mountain corridors and into lowland habitats was likely driven by alternating range expansion during cooling periods and retreat to higher elevations during warming periods [@Hines2008-er; @Stewart2010-jn; @Martinet2018-rb]. Today, they remain characteristically cold-adapted organisms [@Heinrich1994-cv], with peak abundance and diversity in mountain ranges and northern latitudes [@Williams1998-xi]. In these habitats, they are often the dominant guild of flower-visiting insects and the principal pollinators of entomophilous flora [@Goulson2010-lo]. As in the past, so now are mountain habitats critical theaters of ecological change for bumble bees. Warming climates are expected to cause habitat loss and potentially novel competitive pressures [@Kerr2015-dg; @Soroye2020-zb; though see @Guzman2021-my]. At the same time, socioeconomic drivers are causing succession to forest in pastures on mountain slopes and the intensification of agriculture in lowlands, both of which entail a loss of floral resources for bumble bees [@Tasser2002-om; @Tappeiner2006-jp; @Tasser2007-qo; @Walcher2017-kw].

In this paper, we present a study based on ~13,000 interactions between bumble bees (Hymenoptera: *Bombus*) and their floral hosts recorded along a 1400 m elevation gradient, complemented by corresponding surveys of floral abundance. Using this dataset, we investigate elevational responses of richness, abundance, and $\beta$-diversity, both at the level of species (bumble bees and flowering plants, respectively) and at the level of interactions (bumble- bee-flower pairs). Our results reveal nonlinear and, in some cases, contrasting elevational patterns, underscoring the fact that mountain slopes are complex ecological systems where multiple processes interact to produce biological responses. These patterns shed light on the basic ecology of the mountain mutualism between bumble bees and wildflowers, and they suggest how this mutualism might be reshaped by ongoing processes of climate and land use change.

@@

Sampling in each year of the study consisted of repeated visits to each site at approximately weekly intervals. Samples were only collected during periods without rain and when the air temperature was at least 6°C. We divided each 60 x 60 m study plot into 10 evenly-spaced 60 x 6 m transects. During each visit, the observer walked each transect in a zig-zag pattern for five minutes and recorded all interactions between bumble bees and flowers; thus, each sampling event consisted of a total of 50 minutes of observation. Bumblebees observed on a given flower were counted as floral visitors. Bumble bee queens were identified to species level in the field, while workers and males were stored in individually labeled tubes in the freezer for later identification in the laboratory according to [@Amiet1996-jz]. Floral visitation by male bumble bees was recorded during visitation sampling, but we chose to analyze only visitation by queens and workers. Males are often found resting on flowers, particularly late in the year, so their presence on a flower cannot necessarily be interpreted as a foraging visit. In conjunction with bumble bee observations, we estimated the flower cover of each herbaceous or shrubby plant species within the study plot to the nearest 0.01 m^2^; this was done by exhaustive visual inspection of the study plot rather than timed transect walks. To avoid observer bias, floral cover estimates were always performed by the same investigator (K. Kallnik). Plant species were identified following @Lauber2007-ha and @Oberndorfer2001-ro.

@@

We quantified bumble bee abundance as the total number of recorded floral visits per bumble bee species per site-date. We treated the cryptic *Bombus terrestris/lucorum* species pair as a single morphospecies, since these two species cannot be visually distinguished, and we pooled representatives of the parasitic *Psithyrus* subgenus --- *Bombus barbutellus*, *B. bohemicus*, *B. campestris*, *B. flavidus*, *B. quadricolor*, and *B. sylvestris* --- into the species group *B. psithryus*.  *Bombus hypnorum*, *B. humilis*, and *B. lapidarius* were extremely rare in our samples, so we omitted them from abundance analyses but retained them for $\beta$-diversity analyses.

@@

To facilitate clear interpretation with respect to our focal questions, the models above focus on elevational patterns of abundance by summarizing across seasonal time. For those interested in within-season abundance patterns and their interaction with elevation, more complex tensor-product GAMs are included in Appendix S1.

### Species and interaction $\beta$-diversity

For our analyses of species and interaction $\beta$-diversity, we pooled all samples by site irrespective of year. After pooling, we calculated $\beta$-diversity between all pairs of sites for bumble bees, flowering plants, and theirs interaction, respectively, using the binary Jaccard index. We then partitioned total interaction $\beta$-diversity into its components of species turnover (changes in species composition) and interaction rewiring (interaction turnover despite species constancy) [@Novotny2009-my; @Poisot2012-fk].

As an initial analysis, we visualized the relationship between site-wise elevation difference and each metric of $\beta$-diversity, fitting each response with a binomial regression smooth. To investigate the relationship between elevation and $\beta$-diversity more deeply, we performed a second analysis using generalized dissimilarity modeling (GDM) [@Ferrier2007-ji] to analyze the elevational variation in the $\beta$-diversity of bumble bees, flora, and their (unpartitioned) interactions, respectively. In addition to estimating the relationship between differences in a response variable and differences in a predictor variable, GDM captures the slope of this relationship over the range of the predictor variable, revealing potential variation in the amount of change in the response induced by a given change in the predictor [@Ferrier2007-ji]. This enabled us to ask whether the relationship between $\beta$-diversity and elevation exhibits thresholds or other nonlinearities. To control for potential confounding effects of sampling intensity, sampling days per site was included as a covariate. We also included the geographic distances between sites as a covariate to account for potential spatial autocorrelation. The importance of the variables in each model were assessed by a permutation test [@Fitzpatrick-2021aa].

@@

All analyses were conducted in `R` [@RCoreTeam-2021aa]. Data handling and visualization were performed with the `tidyverse` suite [@Wickham2019-aa]. GAM analyses were performed with packages `mgcv` [@Wood_2017], `mgcViz` [@Fasiolo2018aa], and `gratia` [@Simpson2021-aa]. $\beta$-diversity was calculated with the package `vegan` [@Oksanen_2019] and visualized using the packages `ggplot2` [@Wickham2016-aa] and `ggvegan` [@Simpson2019aa]. Calculation and partitioning of interaction $\beta$-diversity were performed with the package `bipartite` [@Dormann2008-aa; Dormann2009-aa]. GDM analysis of $\beta$-diversity was performed with the package `gdm` [@Fitzpatrick-2021aa]. The full workflows of our analyses are available in Appendices S1 and S2, and the data are archived on Dryad: https://datadryad.org/stash/share/M_4Yt9VVXTRm36FSfwwGzUs2OTXP9Fv52jjETdSu6ys.

@@

## Bee-flower interactions and floral survey data

@@

The response of bumble bee richness to elevation exhibited an approximately linear increase up to 1600-1700 m and then leveled off and declined slightly at the upper limit of the gradient **(Figure 3A)**. The year 2012 had a negative effect (p = 0.04) on bumble bee richness relative to the reference year 2010 **(Figure 3B)**, but neither mowing nor abandonment had a significant effect on bumble bee richness relative to the reference management level of grazing (p > 0.05)  **(Figure 3C; Table 1)** In contrast, the response of floral richness exhibited a clear unimodal pattern, peaking between 1300-1500 m **(Figure 3D)**. The years 2011 (p < 0.001) and 2012 (p < 0.03) had a negative effects on floral richness relative to the reference year 2010 **(Figure 3E)**, but again management had no significant effect (p > 0.1) **(Figure 3F; Table 1)**. The response of interaction richness to elevation was intermediate to that of bumble bee and floral richness, peaking around 1500 m and leveling off at higher elevation **(Figure 3G)**. The effect of year 2011 on interaction richness was positive (p < 0.05) and that of year 2012 negative (p < 0.001) **(Figure 3H)**, but management did not have a significant effect (p > 0.1) **(Figure 3I; Table 1)**.

Cumulative bumble bee abundance did not respond significantly to elevation (p > 0.1) **(Figure 4A)** or management (p > 0.1) **(Figure 4C)**, but the year 2011 had a positive effect (p = 0.03) on bumble bee abundance relative to the reference year of 2010 **(Figure 4B; Table 2)**. Floral abundance, however, peaked around 1000 and exhibited an approximately quadratic decline with increasing elevation (p = 0.003) **(Figure 4D)**. The effect of year 2012 was negative (p = 0.002) relative to the reference year 2010 (p = 0.02) **(Figure 4E)**, and the effect of abandoned meadows on floral abundance was positive compared to the reference level of grazed meadows **(Figure 4F; Table 2)**.

@@

## Species and interaction $\beta$-diversity

Floral $\beta$-diversity among our study sites was very high overall and responded steeply to elevation difference between sites **(Figure 3A)**. Even sites at similar elevation exhibited ~40% species turnover, and the most vertically distant sites (elevation difference > 1250 m) differed by more than 85%. Bumble bee $\beta$-diversity was, in comparison to floral $\beta$-diversity, both lower overall and less responsive to elevational difference between sites **(Figure 3A)**. Sites at similar elevation exhibited ~25% species turnover, and species turnover between the most vertically distant sites remained less than 50%.

Total interaction $\beta$-diversity was >75% between sites at similar elevation and approached perfect dissimilarity in the most vertically distant sites **(Figure 3B)**. Partitioning revealed that total interaction $\beta$-diversity was driven primarily by species turnover, which accounted for ~50% of total $\beta$-diversity for sites at similar elevation and ~90% of total $\beta$-diversity for the most vertically distant sites. Species turnover consisted mainly of turnover in the floral community **(Figure 3C)**, though the joint turnover of plants and bumble bees accounted for ~25% of total species turnover between the most vertically distant sites. Interaction rewiring was most significant between sites at similar elevation, where it accounted for ~30% of total $\beta$-diversity, but its share of total $\beta$-diversity declined to ~10% in the most vertically distant sites **(Figure 3B)**.

Bumble bees, flora, and their interactions exhibited marked differences in their rates of $\beta$-diversity in response to elevation. **(Figure 4)**. Bumble bee turnover was relatively high from ~600-1100 m, briefly leveled off, and then accelerated again around 1250 m. Floral Turnover was high from ~600-1000 m, gradually leveled off from ~1000-1500 m, and then accelerated again around the tree line ecotone at ~1500 m. Interaction turnover mirrored the high rates of bumble bee and floral $\beta$-diversity from ~600-1000 m, but it exhibited a markedly different pattern above 1000 m, with relatively low turnover between 1000-1750 m and then a sharp acceleration from ~1750-2000 m. The conditional effect of elevation was significant (p < 0.001) in all models, and its importance was 4-20 time higher that of sampling intensity and geographic distance **(Table 4)**. Nevertheless, geographic distance was a significant predictor in all models (p < 0.001), and sampling intensity was a significant predictor in the bumble bee (p = 0.004) and interaction models (p < 0.001). Our GDMs explained 31% of total deviance in bumble bee $\beta$-diversity, 73% of floral $\beta$-diversity, and 71% of interactions $\beta$-diversity.

@@

Our study revealed strong and sometimes contrasting patterns of elevational structure in the richness, abundance, and $\beta$-diversity of mountain bumble bees and wildflowers. By extending our analysis to the level of species interactions, we show how the elevational patterns within bumble bee and wildflower communities are propagated to the functional relationships between these two guilds. These patterns suggest underlying processes governing the assembly of bumble- bee-wildflower networks in mountain habitats, shedding light on how this mutualism has evolved in the past and how it may respond to the nascent forces of climate and land use change.

@@

The relative stability of the bumble bee community in our study stands in stark contrast to the extreme dynamism of the floral community, as illustrated by the dominance of total interaction $\beta$-diversity by floral species turnover. This pattern is consistent with the findings of @Simanonok2014-ou in the Rocky Mountains of North America, and it is perhaps unsurprising given the exceptional cold-hardiness of bumble bees and the sensitivity of plants to temperature gradients. Nevertheless, it highlights the behavioral challenge that bumble bees have presumably faced for the entirety of their evolutionary history as mountain florivores: how to forage efficiently in a diverse floral community that exhibits rapid spatial and temporal turnover. A fascinating question that could not be answered by our sampling approach is to what extent mountain bumble bees actively exploit the elevational sensitivity of floral communities by adaptively foraging up- or downslope, as suggested by @Lundberg1980-qk. Bumble bees have large foraging ranges and have been shown both to track resources through space and time [@Devoto2014-ay] and to cross forest matrix to reach patches of foraging habitat [@Mola2020-yc]. Assuming an average slope of 20° and a foraging range of 1 km, a bumble bee could travel up- or downslope by more than 340 m, thus spanning an elevation band nearly 700 m wide. Such 3-dimensional foraging would enable bumble bees both to exploit the elevational turnover of floral species and to track preferred species through their elevationally staggered phenology, the latter constituting a sort of physiological time-travel [@Van_Straalen1983-df].

The roughly sigmoidal pattern of bumble bee $\beta$-diversity rate in response to elevation, revealed by our GDM analysis, probably reflects both the reciprocal abundance patterns of high- and low-elevation species and the occurrence of certain rarer species only at middle-elevation sites. As discussed above, though, bumble bee turnover was low over all, demonstrating the robustness of bumble bees to the environmental variation encompassed by our elevation gradient. Floral $\beta$-diversity rate also exhibited a sigmoidal pattern, and its inflection point coincides with the tree line ecotone at ~1500 m. The floristic importance of the tree line has been noted in previous work [e.g. @Pellissier2010-kg; @Descombes2017-id], and our study corroborates these findings specifically for the subset of flora visited by bumble bees. Interestingly, the tree line did not constitute a discrete transition to a new equilibrium of species composition, but rather a threshold marking the beginning of a process of $\beta$-diversity acceleration that continued for the remainder of the elevation gradient. While the most obvious effect of the tree line may be the relatively discrete transition from forest to scrub/grassland, the tree line also initiates a process of angiosperm species attrition that proceeds at a rate of about 40 species per 100 m [@Korner2003-sc], and the accelerated $\beta$-diversity we observed above the tree line mirrors simultaneous patterns of decreasing floral richness and abundance **(Figures 3B, 4B)**.

In contrast, the rate of interaction $\beta$-diversity exhibited a late and very pronounced acceleration at ~1800 m, indicating that the two highest sites at 1984 m and 2032 m differed strongly in interaction composition from the sites ~200 m below them. At these highest sites, the growing season is very short and vegetation is sparse, giving way to bare rock and scree landscapes, and it may be that the last 200 m interval of our elevation gradient imposes the most stringent environmental filter. It is interesting that interaction $\beta$-diversity did not exhibit the inflection point at the tree line ecotone (~1500 m) that we saw for floral $\beta$-diversity, despite the fact that overall interaction $\beta$-diversity was driven primarily by the turnover of the floral community. A possible explanation for this pattern is that there could be an elevational lag in bumble bee floral selection such that most bumble bee species do not switch to supra-tree-line floral species when these species first appear, but rather maintain associations with flora whose ranges span the tree line until the extreme species turnover between 1800 and 2000 m forces them to adopt new floral hosts.
